---
title: "Google Data Analytics Capstone project"
subtitle: "Cyclistic Case Study"
author: "Ramnikhil Palasetty"
date: "27/09/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

### Introdution

I've decided to work with the Cyclistic bike share data for the Google Data Analytics certificate capstone project. I will carry out real-world duties of a junior data analyst for the marketing team at Cyclistic, a made-up bike-share business in Chicago, for the case study.
I will use the six steps of the data analysis process—ask, prepare, process, analyse, share, and act—to provide answers to important business issues.


### The scenario

The future success of Cyclistic, according to Lily Moreno, head of marketing, hinges on increasing the number of annual subscribers. The marketing analyst team is hence interested in seeing how annual members and casual riders utilise Cyclistic bikes differently. The analytics team may be able to create a new marketing plan to turn casual riders into annual members based on these insights.

The following three issues will direct the upcoming marketing campaign:

1. How do yearly members and occasional users of bicycles use them differently?

2. What motivates casual riders to purchase annual Cyclistic memberships?

3. How can Cyclistic utilise social media to persuade casual riders to join?

The first query was given to me by Moreno.


### The Ask phase

* A description of the professional task:

Annual members are significantly more profitable than casual cyclists, according to Cyclistic. So, we aim to create marketing plans and a campaign that aid in turning occasional riders into yearly subscribers.

* My main constituencies are:

Lily Moreno is my manager and the director of marketing. This plan was started by Moreno. the first recipient of the delivery.

2-The executive team: The executive team must approve our recommendations for Moreno's concept to succeed, thus they must be supported with compelling data insights and expert data visualisations.


### The Prepare phase

a Data source
The original bike sharing data set's last 12 months, from January 1, 2021, to December 31, 2021, were extracted and stored as 12 zipped.csv [files] (https://divvy-tripdata.s3.amazonaws.com/index.html). Motivate International Inc. provides the data and grants the licence described in this [license](https://ride.divvybikes.com/data-license-agreement).

Organization & description of the data:

The file naming format is YYYY MM.

Type of file: CSV format

File Content: Each csv file has 13 columns, each of which contains data about the rider type, ride id, start and end times, and location. Row counts in various Excel files range from 49k to 531k.


Data credibility: 

The data set is accurate, full, and dependable for the selected time window.

The information is first-rate and original.

The data set has all the data necessary to provide a complete response to the query.

As rider data from the previous 12 months was used, the data is current.

The Chicago Department of Transportation cited and verified the information.


Data security: Tokenization is used to conceal the personally identifying information of riders.

Backup copies of original files are kept in a different folder.


Data Restrictions: Because riders' personally identifying information is concealed, pass purchases cannot be linked to credit card numbers in order to identify whether frequent riders reside in the Cyclistic service region or whether they have bought many single passes.

### The Process Phase

I cleaned and verified my data using R. Reasons:
More than 5 million rows of data will be included across the 12 data sets combined. There are a maximum of 1,048,576 rows in an Excel worksheet. Moreover, due to issues with file size, some csv files could not be uploaded to BigQuery. R is thus utilised for all activities, including data organisation, cleaning, analysis, and visualisation.


### The Analyze Phase
#### Setting up the environment

Here, I use several libraries that help reading, cleaning, organizing and analyzing the data.

```{r libraries, echo=TRUE, eval=TRUE}
library(tidyverse)
library(janitor)
library(skimr)
library(here)
library(hablar)
library(readxl)
library(data.table)
library(chron)
library(readr)
library(lubridate)
library(magrittr)
library(DescTools)
library(metR)
```

#### Importing data 
Cyclist data from 01/2021 until 12/2021 is imported and read as csv. files. 

```{r csv data collection, echo=TRUE, eval=FALSE}
data_01 <- read_csv("case study/raw data/2021_01.csv")
data_02 <- read_csv("case study/raw data/2021_02.csv")
data_03 <- read_csv("case study/raw data/2021_03.csv")
data_04 <- read_csv("case study/raw data/2021_04.csv")
data_05 <- read_csv("case study/raw data/2021_05.csv")
data_06 <- read_csv("case study/raw data/2021_06.csv")
data_07 <- read_csv("case study/raw data/2021_07.csv")
data_08 <- read_csv("case study/raw data/2021_08.csv")
data_09 <- read_csv("case study/raw data/2021_09.csv")
data_10 <- read_csv("case study/raw data/2021_10.csv")
data_11 <- read_csv("case study/raw data/2021_11.csv")
data_12 <- read_csv("case study/raw data/2021_12.csv")
```

Compare column names each of the files. While the names don't have to be in the same order, they DO need to match perfectly before they are  joined  into one file.

```{r colname inspection, echo=TRUE, eval=TRUE}
colnames(data_01)
colnames(data_02)
colnames(data_04)
colnames(data_05)
colnames(data_06)
colnames(data_07)
colnames(data_08)
colnames(data_09)
colnames(data_10)
colnames(data_11)
colnames(data_12)
```

Afterwards, the 12 data sets are combined into one big data frame of 5595028 observations.

```{r stacking the datasets , echo=TRUE, eval=TRUE}
all_trips <- bind_rows(data_01, data_02, data_03, data_04, data_05, 
                      data_06, data_07, data_08, data_09, data_10, data_11, data_12)
```

####  Clean up and organize data to prepare for analysis

Inspect the new table that has been created.

```{r all_trips inspection, echo=TRUE, eval=TRUE}
colnames(all_trips)  #List of column names
dim(all_trips)  #Dimensions of the data frame
head(all_trips)  #See the first 6 rows of data frame
str(all_trips)  #See list of columns and data types (numeric, character, etc)
```

Then,  columns that list the date, month, day, day_of_week and year of each ride are added. days of the week are assigned the numbers 1:Monday, 2:Tuesday, etc.
This will allow the aggregation of the data by each day, month or day_of_week.

```{r separting ride date and extracting date data, echo=TRUE, eval=TRUE}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%u") #"%A" would deliver names of weekdays
```


Add a "ride_length" calculation to all_trips in seconds and in minutes (2 new columns).

```{r calulate the ride_length in secs, echo=TRUE, eval=TRUE}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
all_trips$ride_length_m <- (difftime(all_trips$ended_at,all_trips$started_at))/60

```


Inspect the structure of the newly added columns.

```{r data inspection, echo=TRUE, eval=TRUE}
str(all_trips)
```


Convert c(ride_length, ride_length_m, day and month) to numeric so that calculation can be executed.

```{r converting variables to numeric, echo=TRUE, eval=TRUE}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
all_trips$ride_length_m <- as.numeric(as.character(all_trips$ride_length_m))
all_trips$month <- as.numeric(all_trips$month)
all_trips$day <- as.numeric(all_trips$day)
is.numeric(all_trips$ride_length)
is.numeric(all_trips$ride_length_m)
is.numeric(all_trips$month)
is.numeric(all_trips$day)
```


After converting and inspecting data, it was noticed that col:ride_length has some negative values, probably because start_time and end_time were swapped for these rides, or the system simply registered and recorded the rides incorrectly. So, negative-seconds rides must be excluded.

```{r droppig rows, echo=TRUE, eval=TRUE}
all_trips_v1 <- all_trips[!( all_trips$ride_length < 0),]
```


#### First analysis step: descriptive analysis on ride length

First of all, a descriptive analysis on ride length [min] is performed.

```{r data statistical summary, echo=TRUE, eval=TRUE, error=TRUE}
all_trips_v1 %>% 
  summarise(max(ride_length_m),min(ride_length_m),mean(ride_length_m)) 
```

The overall average ride length is 21.9 minutes. 


Second, the mode of weekday is calculated, namely which weekday occurs the most or on which weekday most bike are rented. 

```{r plot1: mode of the week_day, echo=TRUE, eval=TRUE, error=TRUE}
  all_trips_v1 %>% 
  group_by(day_of_week) %>% 
  summarise(number_of_rides = n()) %>% 
  ggplot(mapping = aes(x = day_of_week, y = number_of_rides)) + geom_col()
```

The plot shows that most rides were started on Saturday(991047 rides), Saturday(857285 rides) followed by Friday(810508 rides). So most bikes are rented on the weekend. 


Next, a plot of the ride_length or average_duration in minutes for every day of the week for members and casual riders is shown. 

```{r plot2: ride_length per day per rider type, echo=TRUE, eval=TRUE, error=TRUE}
all_trips_v1 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length_m)) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

The plot demonstrates that casual riders rent bikes for longer durations, especially on Sunday, Saturday,  Friday (weekend) and on Monday. Members show a steady riding/using behavior, plus they also tend to ride a little longer on the weekend.  


Here, number of rides per day for every rider type is plotted. 

```{r plot3: number_of_rides per day per rider type, echo=TRUE, eval=TRUE, error=TRUE}
all_trips_v1 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length_m)) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

Surprisingly, and in contrast to the former plot, members begin more rides and thus have higher number of rides on every day of the week except for Saturday and Sunday. 


#### Second analysis step: average ride length depending on rider type and number of each rider type


```{r plot4: mean ride lenght, echo=TRUE, eval=TRUE, error=TRUE}
all_trips_v1 %>%
  group_by(member_casual) %>%
  summarise(max(ride_length_m), min(ride_length_m),avg_ride_length = mean(ride_length_m)) %>% 
  ggplot(aes(x = member_casual, y = avg_ride_length,fill=member_casual)) +
  geom_col()+ scale_y_continuous(breaks = seq(0, 40, by = 5))
```

So, the result shows that casual riders tend to rent bikes for longer mean durations than members (32 min to 13.6 min), in accordance with plot 2. Members probably use bikes to commute, whereas casual riders maybe, among other things, exercising, visiting the city or attending special events.


Here, the overall rider count based on rider type is plotted

```{r plot5: overall rider count by rider type, echo=TRUE, eval=TRUE, error=TRUE}
all_trips_v1 %>%
  group_by(member_casual) %>%
  summarise(rider_count = n()) %>% 
  ggplot(aes(x = member_casual, y = rider_count,fill=member_casual )) +
  geom_col()
```

The plot indicates that more than half of all riders are member riders (3/(3+2.5)=0,55)


#### Third analysis step: Exploring  effect of  seasonality 

Here, the Function "season" of the library "metR" was used to assign season to months:

DJF:winter

MAM:Spring

JJA:Summer

SON:Fall


```{r assigning season to months} 
all_trips_v1$season <- season(all_trips_v1$month)
```

First, let us consider number of rides and ride length by weekday on each season

```{r plot6: number of rides by week_day and rider type and season, echo=TRUE, eval=TRUE, error=TRUE}
all_trips_v1%>%
  group_by(season, day_of_week, member_casual) %>%   
  summarise(number_of_rides = n()						 
  ,avg_ride_length = mean(ride_length_m)) %>% 
ggplot() + geom_col(mapping = aes(x = day_of_week, y = number_of_rides, fill = member_casual), position = "dodge") + facet_wrap(~season) + scale_y_continuous(breaks = seq(0, 400000, by = 50000))
```

```{r plot7: ride_lenght by week_day and rider type and season, echo=TRUE, eval=TRUE, error=TRUE}
all_trips_v1%>%
  group_by(season, day_of_week, member_casual) %>%   
  summarise(number_of_rides = n()						 
  ,avg_ride_length = mean(ride_length_m)) %>% 
ggplot() + geom_col(mapping = aes(x = day_of_week, y = avg_ride_length, fill = member_casual), position = "dodge") + facet_wrap(~season) + scale_y_continuous(breaks = seq(0, 50, by = 10))
```

Plot 6 tells us that number of rides of members is always higher than that of casual riders on every  work day in every season. Weekends are still the time where casual riders bike more than members. The only exception of this trend is in the winter months (Dec, Jan, Feb).
Plot 7 demonstrates that the member group has all year long the average ride length of about 13.6 minutes. Casual riders use bikes  about half an hour long on all days in spring and summer. In winter and fall, the average ride lengths becomes less than 30 minutes. 


Lastly, let us generate a line plot for continuous change of number of rides along the whole year for the two rider types.

```{r plot8: number of rides along the whole year}
all_trips_v1%>%
  group_by(month, member_casual) %>%   
  summarise(number_of_rides = n()						 
  ,avg_ride_length = mean(ride_length_m)) %>% 
ggplot() + geom_line(mapping = aes(x = month, y = number_of_rides, color = member_casual)) + scale_x_continuous(breaks = seq(1, 12, by = 1))
```

The plot indicates, for casual riders/members, ridership peaked around July/August (Summer months being the turning point) and hit the lowest at February before rebounding up swiftly and continuously. 



### The Share phase 

#### Conclusions/Summary of insights 

The amount of time, frequency, and days of the week on which each group peaks varies between members and casual riders:

* Weekends are busiest for casual riding (plot3). It's very likely that they are city visitors who are taking in the sights, or that they are regular Chicago residents who are enjoying a weekend bike ride. This is supported by the fact that the average ride time for casual riders (plot2) is longer and increases on weekends.

* Members tend to ride for shorter periods of time than casual riders. This may be explained by saying that the majority of the members commute to work on bicycles. This clarification would also shed light on why members only ride for little periods of time. They travel from point A to point B, which is essentially the same distance and ride length every time.

* Beginning in February (from Spring through Summer), ridership begins to increase. It then begins to decline in August (plot 8). (from Fall through winter). The seasonal fluctuations are what cause this association. More individuals begin to cycle when the weather begins to warm up and become more pleasant in February (the beginning of Spring), and the opposite is true as the weather turns cooler and less pleasant around September (start of Fall).

* More than 50% of the riders (plot 5) are annual members, indicating that the business has already attained a certain level of customer loyalty. This suggests a good message, namely that the company will be able to persuade many casual riders to become members and to maintain the satisfaction of the new members.


#### Recommendations

When you sign up for a membership, you get discounts for longer rides.
Longer rides may be eligible for a rewards programme once they join.

* The marketing initiative should be started between February and August, when the volume of journeys taken by casual riders peaks.

* The marketing effort can feature weekend-only membership at a fair price since casual rider usage peaks on the weekends. This might entice infrequent riders to become subscribers.

* A ride-length-based tariff plan might be included in the campaign (perhaps only on weekends): More bicycles, less money!
This gives the member rides more motivation to cycle farther.

* As an alternative, lengthier rides may be rewarded with advantages like gift cards for discounts.














